Renormalization

### 2014-05-19 papers/renormalization/paper.tex
add blurb on RGT to (still mostly empty) RG paper Tested on bennet. Took 0:21.17.

\fixme{Here we will explain the RGT that we use, and how it works
  out.}

>>>>

\subsubsection{Defining our approximation}
Let's begin by considering the partition function and free energy, as
computed in the canonical ensemble:
\begin{align}
  Z &= \sum_i^{\scriptscriptstyle\begin{array}{c}\text{\tiny all}\\\text{\tiny microstates}\end{array}}
       e^{-\beta E_i}, & F &= -kT\log Z
\end{align}
We now consider a ``smooth'' configuration that contains only
microstates that are periodic with a given periodicity $2^j\lambda$:
\begin{align}
  Z_j &= \sum_i^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny microstates}\\
                  \text{\tiny periodic }2^j\lambda\\
                 \end{array}}
       e^{-\beta E_i}, & F_j &= -kT\log Z_j
       \label{eq:Zn}
\end{align}
This defines a family of free energies that contain fluctuations with
a variety of length scales.  For the moment we will assume a simple
cubic periodicity, although it may be advantageous to select face
centered cubic (FCC) instead.  So far, we have defined a family of
free energy functions (each containing longer length-scale
fluctuations) that we can compute numerically using Monte Carlo, but
we have not yet constructed a means to approximate these functions.
Before developing this approximation, let's reorganize this sum over
periodic states.
\begin{align}
  Z_j(N) &= \sum_{N_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       \sum_{i_1\cdots i_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny microstates}\\
                  \text{\tiny within octants}\\
                 \end{array}}
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
\end{align}
where here $g_{N_1\cdots N_8}$ is a multiplicity factor
corresponding to the number of ways the cell can be divided into
subcells with occupancies of $N_1\cdots N_8$.  This is so far only a
different way of writing Eq.~\ref{eq:Zn}.  At this point, we can
multiply and divide by a few factors, with a goal of replacing the
unknown $E_{i_1\cdots i_8}^{(j)}$ with something that is easier to
approximate.
\begin{multline}
  F_j(n) = -kT\log\Bigg(
    \sum_{N_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       \\
       \sum_{i_1\cdots i_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny microstates}\\
                  \text{\tiny within octants}\\
                 \end{array}}
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
  \Bigg)
\end{multline}
Now we consider the same free energy computed without accounting for
the energy in the fluctuations:
\begin{multline}
  F_{j-1}(n) = -kT\log\Bigg(
    \sum_{N_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       \\
       \sum_{i_1\cdots i_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny microstates}\\
                  \text{\tiny within octants}\\
                 \end{array}}
       e^{-\beta \left(E_{i_1}^{(j-1)} + \cdots + E_{i_8}^{(j-1)}\right)}
  \Bigg)
\end{multline}
We can now recognize that each of the octants is just a similar smooth
system.
\begin{multline}
  F_{j-1}(n) = -kT\log\Bigg(
    \sum_{N_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       \\
       e^{-\beta \left( F_{j-1}(N_1) + \cdots F_{j-1}(N_8) \right)}
  \Bigg)
\end{multline}
Here I assume that the fluctuations will only be important in so far
as the affect the free energy directly.  Putting these last two
equations together, we can find a formula for the energy difference as
we include the larger fluctuations:
\begin{multline}
  F_j(n) - F_{j-1}(n) =\\
  -kT\log\left(\frac{
    \sum_{AAN_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       \sum_{i_1\cdots i_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny microstates}\\
                  \text{\tiny within octants}\\
                 \end{array}}
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
  }{
    \sum_{N_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
        e^{-\beta \left( F_{j-1}(N_1) + \cdots F_{j-1}(N_8) \right)}
  }
  \right)
\end{multline}
At this point, we recognize that we can put anything we like in the
numerator \emph{and} the denominator, which will help us to prevent
the exponentials from getting too hairy.  This is really just a shift
of energy for both the numerator and denominator partition functions.
\begin{multline}
  F_j(n) - F_{j-1}(n) =\\
  -kT\log\left(\frac{
    \sum_{NN_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       \sum_{i_1\cdots i_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny microstates}\\
                  \text{\tiny within octants}\\
                 \end{array}}
       e^{-\beta \left(E_{i_1\cdots i_8}^{(j)}  - F_{j-1}(n) \right)}
  }{
    \sum_{BN_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       e^{-\beta \left( F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) \right)}
  }
  \right)
\end{multline}
At this stage we have still not introduced an uncontrolled
approximation.  Each $F_{j}$ is exactly defined, and the exact free
energy is found by taking a limit $j\rightarrow \infty$.  \fixme{I
  need to clarify the scale of each $F$.  Some of the above are about
  systems $\frac18$ the volume of others.}

The flavor of the approximation used by White and others in GRG is now
clarified.  We begin by assuming that we already have an adequate (and
tractable) approximation for $F_{j-1}$.  Thus the denominator
partition function is readily computable.  We further simplify the
numerator by multiplying and dividing by the same quantity, and
simplify the notation by defining
\begin{align}
  F_j(N_1\cdots N_8) &= \sum_{i_1\cdots i_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny microstates}\\
                  \text{\tiny within octants}\\
                 \end{array}}
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
\end{align}
and thus rewrite the energy difference as
\begin{multline}
  F_j(n) - F_{j-1}(n) =\\
  -kT\log\left(\frac{
    \sum_{CN_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       e^{-\beta \big(F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) +
           \big(F_j(N_1\cdots N_8) - F_{j-1}(N_1) - \cdots - F_{j-1}(N_8)\big) \big)}
  }{
    \sum_{N_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       e^{-\beta \left( F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) \right)}
  }
  \right)
\end{multline}
At this point we can recognize and identify two energy differences as
\begin{align}
  \bar{F}(N_1\cdots N_8) &= F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) \\
  \bar{U}(N_1\cdots N_8) &= F_j(N_1\cdots N_8)
  - \left(F_{j-1}(N_1) + \cdots + F_{j-1}(N_8) \right)
\end{align}
and using these definitions can rewrite our energy difference in a
more compact form:
\begin{multline}
  F_j(n) - F_{j-1}(n) =\\
  -kT\log\left(\frac{
    \sum_{CN_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       e^{-\beta \big( \bar{F}(N_1\cdots N_8) + \bar{U}(N_1\cdots N_8)
         \big)}
  }{
    \sum_{N_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
                  \text{\tiny occupancies}\\
                  \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}
       e^{-\beta \bar{F}(N_1\cdots N_8)}
  }
  \right)
\end{multline}
At this point, we can recognize $\bar{F}(N_1\cdots N_8)$ as the free
energy change due to oscillations \emph{if we neglect the actual
  energy of the oscillations themselves, but instead make a ``local''
  approximation}.  Meanwhile $\bar{U}(N_1\cdots N_8)$ is the free
energy change due to the energy of oscillations.

\fixme{I have now ``defined'' the terms used by Forte in his
  notation of GRG.  On the to-do list: }
\begin{enumerate}
\item \fixme{Establish what $g_{N_1\cdots N_8}$ is.}
\item \fixme{Elaborate on the approximation (ignoring corners) that
  leads from a $\sum_{N_1\cdots N_8}^{\scriptscriptstyle\begin{array}{c}
    \text{\tiny occupancies}\\ \text{\tiny of octants}\\
                 \end{array}}
       g_{N_1\cdots N_8}$ to an integral over density
       amplitudes.}
\item \fixme{Develop an approach for approximating $\bar{U}$.  Perhaps
  is $F_{j-1}$ is expressed as a functional, we can use a non-local
  DFT to approximate $\bar U_j$?  The standard GRG approach seems to
  be to use a simple mean-field or thermodynamic perturbation theory
  approach to approximate $\bar U$, suggesting that a simple
  (standard) DFT might work just fine for this.}
\end{enumerate}
